---
title: "Regression with Slice Sampling"
author: "Andreas Hofheinz"
date: "25 11 2020"
output: pdf_document
---

```{r setup, include=FALSE}
set.seed(20201128)
knitr::opts_chunk$set(echo = TRUE)
```

```{r, warning=FALSE, message=FALSE}
library(bisemSliceSampler)
library(latex2exp)
library(patchwork)
library(ggplot2)
```

```{r, warning=FALSE, message=FALSE}
data(BostonHousing, package  = "mlbench")
keep_cols <- c("crim", "lstat", "age", "medv")
housing_sub <- subset(BostonHousing, select = keep_cols)
```

```{r}
#summary(BostonHousing)
```


```{r}
# Add column with all 1s for intercept
X <- cbind("intercept" = 1, as.matrix(subset(housing_sub, select = -medv)))
y <- housing_sub$medv
```

**Model:**

$y_{\text {medv }, i}=\beta_{0}+\beta_{1} x_{\mathrm{crim}, i}+\beta_{2} x_{\mathrm{lstat}, i}+\beta_{3} x_{\text {age }, i}+\beta_{4} x_{\text {medv }, i}+\epsilon_{i}, \quad i=1, \cdots, n$ \
with $\epsilon_{i} \stackrel{\mathrm{iid}}{\sim} \operatorname{Normal}\left(0, \sigma^{2}\right)$


$\Longrightarrow y_i \mid \boldsymbol{\beta}, \sigma^{2} \stackrel{\mathrm{i.i.d}}{\sim} \operatorname{N}\left({\mathbf{x_i}}^T\beta, \sigma^{2}\right)$

\bigskip


**Likelihood:**

$L\left(\mu, \sigma^{2}\right) \propto \left(\sigma^{2}\right)^{-n / 2} \exp \left(-\frac{1}{2 \sigma^{2}} \sum_{i=1}^{n}\left(y_{i}-{\mathbf{x}_i}^T\beta\right)^{2}\right)$

\bigskip



**Prior**

In this example we use the reference prior for multiple linear regression:

$p\left(\boldsymbol{\beta} , \sigma^{2}\right) \propto \frac{1}{\sigma^{2}}$


\bigskip

**Posterior**

$$
p\left(\boldsymbol{\beta} , \sigma^{2} \mid \boldsymbol{y} \right)
\propto \left(\sigma^{2}\right)^{-n / 2} \exp \left(-\frac{1}{2 \sigma^{2}} \sum_{i=1}^{n}\left(y_{i}-{\mathbf{x}_i}^T\beta\right)^{2}\right) * \frac{1}{\sigma^{2}}
$$


**Log-Posterior**

$$
log(p\left(\boldsymbol{\beta} , \sigma^{2} \mid \boldsymbol{y} \right))
\propto -\frac{n+2}{2} \log \left(\sigma^{2}\right)-\frac{1}{2 \sigma^{2}} \sum_{i=1}^{n}\left(y_{j}-{\mathbf{x}_i}^T\beta\right)^{2}
$$

```{r}
logpost <- function(theta) {
  beta <- theta[-length(theta)]
  sigma_sq <- theta[length(theta)]
  n <- length(y)
  -((n + 2)/2 * log(sigma_sq)) - (1/(2 * sigma_sq)) * crossprod(y - (X %*% beta))
}
```

Note: $\boldsymbol{\theta} = (\boldsymbol{\beta}, \sigma^{2}) = (\beta_0, \beta_1, \beta_2, \beta_3, \sigma^{2})$

\bigskip


**Sampling from Log-Posterior**

```{r}
w_intercept <- 10
w_rest <- 0.4
w <- c(w_intercept, rep(w_rest, times = ncol(X)))
n_samples <- 2000
theta_init <- c(rep(0, ncol(X)), 10)
```


```{r, message=FALSE}
post_samples  <- slice_sampler(
  logpost , x_init = theta_init, w = w, n_samples = n_samples
)
```

\newpage

**Convergence Diagnostic**

```{r, fig.height=8, echo=FALSE}
burn_in <- 300
# Remove samples for sigma_sq
beta_samples <- post_samples[, -ncol(post_samples)]


# Graph Sample Number vs Value
df_beta_samples <- as.data.frame(beta_samples)
names(df_beta_samples) <- c("beta0", "beta1", "beta2", "beta3")
df_beta_samples$samples <- 1:nrow(beta_samples)
df_tidy <- df_beta_samples %>%
  tidyr::pivot_longer(beta0:beta3, names_to = "coef") %>%
  dplyr::mutate(coef = as.factor(coef))

levels(df_tidy$coef) <- c(
  beta0  = TeX("$\\beta_0$"),
  beta1  = TeX("$\\beta_1$"),
  beta2  = TeX("$\\beta_2$"),
  beta3  = TeX("$\\beta_3$")
)

g1 <- ggplot(df_tidy, aes(samples , value)) +
  geom_line() +
  facet_wrap(~coef, ncol = 1, scales = "free_y", labeller = label_parsed) +
  geom_vline(xintercept = burn_in, color = "red", linetype = "dashed") +
  theme_bw()

max_lag <- 200
ls_acfs <- apply(beta_samples, MARGIN = 2, FUN = acf, lag.max = max_lag, plot = FALSE)
acfs <- unlist(lapply(ls_acfs, `[[`, 1))
df_acfs <- data.frame(
  coef = as.factor(sapply(c("beta0", "beta1", "beta2", "beta3"), rep, times = max_lag + 1)),
  lag = 0:max_lag,
  acf = acfs
)

levels(df_acfs$coef) <- c(
  beta0  = TeX("$\\beta_0$"),
  beta1  = TeX("$\\beta_1$"),
  beta2  = TeX("$\\beta_2$"),
  beta3  = TeX("$\\beta_3$")
)

g2 <- ggplot(df_acfs, aes(lag, acf)) +
  geom_bar(stat = "identity") +
  geom_hline(yintercept =  1.96 * n_samples^(-0.5), color = "blue", linetype = "dashed") +
  geom_hline(yintercept = -1.96 * n_samples^(-0.5), color = "blue", linetype = "dashed") +
  facet_wrap(~coef, ncol = 1, scales = "free_y", labeller = label_parsed) +
  theme_bw()

g1 + g2
```


\newpage

**Convergence Diagnostic - 300 burn-in samples removed**

```{r, fig.height=8, echo=FALSE}
# Remove samples for sigma_sq and 300 burn-in
beta_samples_no_burnin <- beta_samples[(burn_in + 1):nrow(beta_samples), ]

# Graph Sample Number vs Value
df <- as.data.frame(beta_samples_no_burnin)
names(df) <- paste0("beta", 0:3)
df$samples <- 1:nrow(beta_samples_no_burnin)
df_tidy <- df %>%
  tidyr::pivot_longer(beta0:beta3, names_to = "coef") %>%
  dplyr::mutate(coef = as.factor(coef))

levels(df_tidy$coef) <- c(
  beta0  = TeX("$\\beta_0$"),
  beta1  = TeX("$\\beta_1$"),
  beta2  = TeX("$\\beta_2$"),
  beta3  = TeX("$\\beta_3$")
)

g1 <- ggplot(df_tidy, aes(samples , value)) +
  geom_line() +
  facet_wrap(~coef, ncol = 1, scales = "free_y", labeller = label_parsed) +
  theme_bw()

max_lag <- 200
ls_acfs <- apply(beta_samples, MARGIN = 2, FUN = acf, lag.max = max_lag, plot = FALSE)
acfs <- unlist(lapply(ls_acfs, `[[`, 1))
df_acfs <- data.frame(
  coef = as.factor(sapply(c("beta0", "beta1", "beta2", "beta3"), rep, times = max_lag + 1)),
  lag = 0:max_lag,
  acf = acfs
)

levels(df_acfs$coef) <- c(
  beta0  = TeX("$\\beta_0$"),
  beta1  = TeX("$\\beta_1$"),
  beta2  = TeX("$\\beta_2$"),
  beta3  = TeX("$\\beta_3$")
)

g2 <- ggplot(df_acfs, aes(lag, acf)) +
  geom_bar(stat = "identity") +
  geom_hline(yintercept =  1.96 * n_samples^(-0.5), color="blue", linetype="dashed") +
  geom_hline(yintercept = -1.96 * n_samples^(-0.5), color="blue", linetype="dashed") +
  facet_wrap(~coef, ncol = 1, scales = "free_y", labeller = label_parsed) +
  theme_bw()

g1 + g2
```


\newpage

**Coefficients and Credibility Intervals**

```{r, echo=FALSE, comment=""}
summary_raw <- t(apply(beta_samples_no_burnin, 2, function(x) {
  c(
    round(mean(x), 3),
    round(quantile(x, prob = 0.025), 3),
    round(quantile(x, prob = 0.975), 3)
  )
}))
summary_df <- data.frame(summary_raw)
rownames(summary_df) <- colnames(X)
colnames(summary_df) <- c("Mean", "2.5%", "97.5%")
summary_df
```

**Comparison with frequentist calculation**

Expected to be similar as non-informative prior was chosen.

```{r, echo=FALSE, comment=""}
mylm <- lm(y ~ X - 1)
summary_mylm <- as.data.frame(cbind(mylm$coefficients, confint(mylm)))
summary_mylm <- round(summary_mylm, 3)
rownames(summary_mylm) <- substring(rownames(summary_mylm), 2)
colnames(summary_mylm) <- c("Mean", "2.5%", "97.5%")
summary_mylm
```



